---
title: "run_analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results= "hide")
```


## Including Code

### run_analysis.R that does the following:
* Merges the training and the test sets to create one data set.
* Extracts only the measurements on the mean and standard deviation for each measurement.
* Uses descriptive activity names to name the activities in the data set
* Appropriately labels the data set with descriptive variable names.
* From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.


## 0 load required libraries and download data for the project. 
```{r}
	rm(list = ls(all.names = TRUE))
	wd <<- "C:/Users/user/Documents/R/3.Getting And Cleansing Data/week4/final"
	gwd <- getwd()
	
	if ( gwd != wd ) {setwd (wd)}
	
	if("data.table" %in% rownames(installed.packages()) == FALSE) {install.packages("data.table")}
	require(data.table)
	if("plyr" %in% rownames(installed.packages()) == FALSE) {install.packages("plyr")}
	require(plyr) 
	if("reshape" %in% rownames(installed.packages()) == FALSE) {install.packages("reshape")}
	require(reshape)
	if("downloader" %in% rownames(installed.packages()) == FALSE) {install.packages("downloader")}
	require(downloader)
	
	###  download data archive and unzip 
	fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
	dataFile = "./getdata-projectfiles-UCI HAR Dataset.zip"
	if(file.exists(dataFile)) {
		message("File is already downloaded")
	} else {
		download(fileUrl, dataFile, mode = "wb")
		unzip(dataFile)
	}
	

```
## 1    Merges the training and the test sets to create one data set.

	### read train files
```{r}
		x_train <- read.table("./UCI HAR Dataset/train/X_train.txt")
		y_train <- read.table("./UCI HAR Dataset/train/y_train.txt")
		subject_train <- read.table("./UCI HAR Dataset/train/subject_train.txt")
	### read test files

		x_test<- read.table("./UCI HAR Dataset/test/X_test.txt")
		y_test <- read.table("./UCI HAR Dataset/test/y_test.txt") 
		subject_test <- read.table("./UCI HAR Dataset/test/subject_test.txt")

			### TRAINING DATA STRUCTURE SUMMARY  
				# summary(x_train)
				# summary(y_train)
				# summary(subject_train)

			### TEST DATA STRUCTURE SUMMARY  
				# summary(x_test)
				# summary(y_test)
				# summary(subject_test)
				
```				
## Uses descriptive activity names to name the activities in the data set
	### read features file 
```{r}
	features <- read.table("./UCI HAR Dataset/features.txt")
	

	### Appropriately labels the data set with descriptive variable names.
		names(x_train)<-features[,2]
		names(x_test)<-features[,2]

		names(y_train)<-c("ActivityId")
		names(y_test)<-c("ActivityId")

		names(subject_train)<-c("SubjectId")
		names(subject_test)<-c("SubjectId")



	### column bind subject, activity and features data:
		train_data <- cbind(subject_train, y_train, x_train)
		test_data  <- cbind(subject_test, y_test, x_test)

	### row bind train and test data:
		mergedData <- rbind(train_data, test_data)
```
## 3 Extracts only the measurements on the mean and standard deviation for each measurement.
	### use pattern matching to extract mean and std data and keep subjectid and activityid
```{r}
				mergedData <- mergedData[,grepl("SubjectId|ActivityId|mean\\(\\)|std\\(\\)", names(mergedData))]
```
## Uses descriptive activity names to name the activities in the mergedData set
```{r}
					### join mergedData with activity_labels on ActivityId
		activity_labels <- read.table("./UCI HAR Dataset/activity_labels.txt")
		names(activity_labels) <- c("ActivityId", "Activity")
		mergedData <- join( mergedData,activity_labels, by="ActivityId",  match="first")	 
	### tidy column names
	### remove "()"
		names(mergedData) <- gsub("\\(\\)", "", names(mergedData)) 
	### replace dashes with undescores 
		names(mergedData) <- gsub("-", "_", names(mergedData))  
	### capitalize Mean and Std
		names(mergedData) <- gsub("mean", "Mean", names(mergedData)) 
		names(mergedData) <- gsub("std", "Std", names(mergedData)) 
	
	###  move activity column in the beginning  &  remove activityId index column 
		col_idx <- grep("^Activity$", names(mergedData))
		tidyData <- mergedData[, c(col_idx, (1:ncol(mergedData))[-col_idx])]
		rm(mergedData)
		tidyData <- tidyData[,-2]
		str(tidyData)
	
	### Write the dataset of activities/subjects/features to a file   
	tidyDataFile <- "GettingCleaningData_Project.TXT"
	write.table(tidyData, tidyDataFile, row.names=FALSE)
```
## 5 From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r}
	### set SubjectId, ActivityId, Activity as a key.
		 tidyData <- as.data.table(tidyData)
		# tidyData <- tbl_df(tidyData)
		 setkey(tidyData, SubjectId, Activity)

	### reshape dataset from a short and wide format to a tall and narrow format.

		tidyData <- data.table(melt(tidyData, key(tidyData), variable.name="featureCode"))

## Create a data set with the average of each featureCode for each Activity and each SubjectId.
	setkey(tidyData, SubjectId, Activity,featureCode)
  
	### Using cast to move featureCode into columns 
	tidyDataSummary <- cast(tidyData, SubjectId+Activity ~ featureCode,mean)
	tidyDataSummary <- as.data.table(tidyDataSummary)
	rm(tidyData)
	str(tidyDataSummary)
    ### Write the final tidy dataset of activities/subjects/features Means to a file   
    tidyDataFile <- "GettingCleaningData_Project_Mean.TXT"
    write.table(tidyDataSummary, tidyDataFile, row.names=FALSE)
```